---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This document is created for Peer Assessment 1 in Reproducible Research course on coursera.org.

## Loading and preprocessing the data
Assuming the file to be processed is in the working directory the following code
will unzip & load the source data into a data frame called *raw_data* for further analyses.

```{r}
unzip(zipfile = "activity.zip")
raw_data <- read.csv(file = "activity.csv")
```

To gain a better understanding of the data, I shooted some summary functions:
```{r}
str(raw_data)
summary(raw_data)
```

## Mean total number of steps taken per day
This part of the assignment, I ignored the missing values in the dataset.
```{r}
raw_data_nona <- raw_data[!is.na(raw_data$steps),]
```
```{r, echo=FALSE}
rowdiff <- nrow(raw_data)-nrow(raw_data_nona)
```
That eliminated **`r rowdiff`** rows from the original dataset.

To prepare a histogram of the total number of steps taken each day I used *dplyr*
package to sum total number of steps for each day included in the observed period.
```{r}
library(dplyr)
tot_steps <- raw_data_nona %>% 
             group_by(date) %>% 
             summarise(total_steps = sum(steps))

hist(tot_steps$total_steps, main="", xlab="Total steps per day")
```

Histogram shows that our population is close to normal distribution.

Additional descriptive statistic measures:

 - Mean: **`r mean(tot_steps$total_steps)`**
 - Median: **`r median(tot_steps$total_steps)`**

## Average daily activity pattern

To discover activity pattern I calculated a mean figure for each 5-minute interval across the entire observation period and placed the result on a line plot.
```{r}
avg_steps_intval <- raw_data_nona %>%
                    group_by(interval) %>%
                    summarise(avg_steps = mean(steps))

plot(x = avg_steps_intval$interval, y = avg_steps_intval$avg_steps, type="l", ylab="Average steps in the given interval", xlab="Measured intervals")
```

This shows us that the following 5-minute interval contains the maximum number of steps on average: **`r arrange(avg_steps_intval, desc(avg_steps))[1,1]`**

## Inputing missing values

Total number of rows with missing values in the dataset: **`r length(raw_data[is.na(raw_data$steps),1])`**

I chose a strategy of filling in all of the missing values that consists of replacing *'NA'* to the mean of that certain time interval. This cleared dataset went into *clear_data* dataframe.

Created a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
col_replaced_NA  <- with(
                        left_join(raw_data,
                                  avg_steps_intval,
                                  by="interval"
                                  ), 
                        ifelse(is.na(steps)==TRUE, 
                               avg_steps, 
                               steps
                               )
                        )

clear_data  <- data_frame(steps = col_replaced_NA, 
                          date = raw_data$date, 
                          interval = raw_data$interval)
```

Calculating distribution of cleared data to raw data:
```{r}
tot_steps_cd <- clear_data %>% 
                group_by(date) %>% 
                summarise(total_steps = sum(steps))

hist(tot_steps_cd$total_steps, xlab="Total steps per day", main = "")
```


## Differences in activity patterns between weekdays and weekends

I used *weekdays()* function to identify whether the given date is either weekday or belong to weekend. The dataset with the filled-in missing values has been used for this part.
Also, I created a new factor variable, called *weekday_weekend*, with two levels ("weekday" or "weekend") and incorporated that in the *clear_data_w* data frame. 

```{r}
clear_data_w <- cbind(clear_data,
                      weekday_weekend = factor(
                                              with(clear_data, 
                                                   ifelse(
                                                          weekdays(as.Date(as.character(clear_data$date))) %in% c("Sunday","Saturday"),
                                                          "weekend", "weekday"
                                                          )
                                                   )
                                               )
                      )
```

I made a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). (I used ggplot2 library for graph creation.)

```{r}
tot_steps_cdw <- clear_data_w %>% 
                group_by(interval, weekday_weekend) %>% 
                summarise(total_steps = sum(steps))


library(ggplot2)

ggplot(tot_steps_cdw, aes(interval, total_steps)) + 
    geom_line() + 
    facet_grid(weekday_weekend ~ .) +
    ylab("Average number of steps") +
    xlab("Measured intervals")
```
