---
title: "Course Project - Regression Models"
output: html_document
---

This document is created for "Course Project" assignment in the framework of Regression Models course (part of Data Science Specialization by Johns Hopkins University Bloomberg School of Public Health on Coursera).

# The Challenge

Working for *Motor Trend*, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon. They are particularly interested in the following two questions:

 - Is an automatic or manual transmission better for MPG?
 - Quantify the MPG difference between automatic and manual transmissions?

# Executive Summary

To be detailed...

# Data Overview

Data to be analysed for this exercise reside as a built-in object, called `mtcars`, comes along with any R distribution as an extract from the 1974 Motor Trend US magazine that comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models) Let's take a first look at it:

```{r message=FALSE, echo=FALSE}
# First, I load all the necessary R packages for further analyses:
library(ggplot2); library(reshape); library(plyr); library(dplyr); library(leaps); library(GGally); #library("printr");
```
 
```{r echo=FALSE}
# Overview of the dataset
mtc <- mtcars; str(mtc)
```

First few rows of the dataset:

```{r echo=FALSE}
head(mtc)
```

Each row represents a model that can be seen in the row names and each column is an attribute of a particular model. Further description is available in R documentation (`?mtcars`), that is presented below:

|Variable|Description|
|--------|-----------|
|mpg|Miles/(US) gallon|
|cyl|Number of cylinders|
|disp|Displacement (cu.in.)|
|hp|Gross horsepower|
|drat|Rear axle ratio|
|wt|Weight (lb/1000)|
|qsec|1/4 mile time|
|vs|V/S|
|am|Transmission (0 = automatic, 1 = manual)|
|gear|Number of forward gears|
|carb|Number of carburetors|

# Exploratory Data Analyses

I start with some initial explaratory data analysis to get a better insight of the potential patterns in the data set. Below I create pairwise scatterplots for every variable.

```{r cache=TRUE, echo=FALSE}
ggpairs(mtc, axisLabels="none", 
        lower = list(continuous = "smooth", params=c(colour="blue")),
        upper=list(params=list(size=3))) + theme(panel.grid.major = element_blank())
```

Let's also take a look at the outcome variable (miles per gallon), how it varies by automatic vs manual transmission.

```{r echo=FALSE}
ggplot(mtc, aes(x=mpg)) + geom_density(fill="black", alpha=0.3, show_guide=TRUE) + 
  geom_density(aes(fill=factor(am, labels = c("auto","manual"))), alpha=0.6) + 
  guides(fill = guide_legend(title = "Transmission type")) +
  labs(x="Miles per gallon",  y="Probability", title="Distribution of Miles Per Gallon")
```

At this point I will form a hypothesis that cars with automatic transmission can travel less distance per gallon then their counterparties with manual transmission. To check whether this pattern happened by random, I performed the necessary statistical analyses in the subsequent chapters.

# Model fitting and inference

First I test if an automatic or manual transmission better for MPG. My initial hypothesis is that cars with automatic transmission consume more fuel, thus have lower range, than cars with manual transmission. 

|Population|Observed Mean|Standard Deviation|
|----------|:-----------:|:----------------:|
|Entire population|`r round(mean(mtc$mpg),2)`|`r round(sd(mtc$mpg),2)`|
|Cars with automatic transmission|`r round(mean(mtc$mpg[mtc$am==0]),2)`|`r round(sd(mtc$mpg[mtc$am==0]),2)`|
|Cars with manual transmission|`r round(mean(mtc$mpg[mtc$am==1]),2)`|`r round(sd(mtc$mpg[mtc$am==1]),2)`|

I fit a linear regression model on the type of transmission (as categorical predictor) having miles per gallon as outcome. The estimations in the table below are in comparison to automatic transmission. 

```{r}
summary(lm(mpg ~ factor(am), mtc))$coef
```

This says that in the group of cars with manual transmission we expect a `r summary(lm(mpg ~ factor(am), mtc))$coef[2,1]` miles per gallon efficiency improvement comparing to group of cars with automatic transmission. Intercept (`r summary(lm(mpg ~ factor(am), mtc))$coef[1,1]`) represents the mean of "mpg" for automatic transmission group. Since the p value for this difference is close to zero, the mean difference between the two groups is statistically significance.

# Model fit / model selection strategy

This part is about selecting the most relevant variables that explain miles per gallon measure most precisely. In order to find the best fitting linear regression model I have revised outcomes of some model selection algorythm.

## Stepwise variable selection

```{r results='hide', echo=FALSE}
full_model <- lm(mpg ~ ., mtc)
final_model <- step(full_model)
```

The first, most obvious try for variable selection is the stepwise algorythm, provided by `step()` function. The following model has been selected in a _backward_ stepwise process as the best fit, that incorporates the following variables: `r names(final_model$coefficients[1:4])` with the corresponding coefficients of `r final_model$coef`.

## Regression subset selection including exhaustive search

To verify the selected variables, I performed a different method for variable selection provided by `leaps` R library. (Figure 3) This result proofed the previously selected three variables.

To be detailed...

# Appendix



## Figure 2: Outcome - Variable Regression Matrix

```{r echo=FALSE}
# Normalization
#mtcars_var_scaled <- data.frame(scale(mtcars)) %>% select(cyl, disp, hp, drat, wt, qsec, vs, am, gear, carb)
#mtcars_scaled <- cbind(mpg=mtcars$mpg, mtcars_var_scaled)
# Melt
#mtcars_sc_m <- melt(mtcars_scaled, id="mpg")
# Plotting
#ggplot(mtcars_sc_m, aes(y=mpg, x=value)) + geom_point(position="jitter") + facet_wrap(~variable, nrow=2)

# Correlation table
# ggpairs(mtcars, lower = list(continuous = "smooth"))
```

# Figure 3: Regression subset selection including exhaustive search

```{r echo=FALSE}
subset <- regsubsets(mpg ~ cyl+disp+hp+drat+wt+qsec+vs+am+gear+carb, mtc)
plot(subset)
```

# Pointing criteria

Todo:

 - interpret the coefficients
 - do some exploratory data analyses
 - fit multiple models and detail their strategy for model selection
 - answer the questions of interest or detail why the question(s) is (are) not answerable
 - do a residual plot and some diagnostics
 - quantify the uncertainty in their conclusions and/or perform an inference correctly

Done:

 - include an executive summary
 - done in Rmd (knitr)
 - brief (about 2 pages long) for the main body of the report and no longer than 5 with supporting appendix of figures

# Obsolete content

```{r message=FALSE, echo=FALSE, eval=FALSE}
mtc$am <- factor(mtc$am); mtc$gear <- factor(mtc$gear); mtc$vs <- factor(mtc$vs);
levels(mtc$am)[levels(mtc$am)=="0"] <- "auto"; levels(mtc$am)[levels(mtc$am)=="1"] <- "man";

```


```{r eval=FALSE}
table(mtcars$am)
am_means <- data.frame(value=rbind(totalmean=mean(mtcars$mpg), automean=mean(mtcars[mtcars$am==1,]$mpg), manmean=mean(mtcars[mtcars$am==0,]$mpg)))
ggplot(mtcars, aes(y=mpg, x=factor(am))) + geom_boxplot()
mtcars_melted <- melt(mtcars, id="mpg")

null <- lm(mpg ~ 1, mtcars)

step(null, scope=list(lower=null, upper=full), direction="forward") 
step(full, data=mtcars, direction="backward") 
step(null, scope = list(upper=full), data=mtcars, direction="both")
```
