---
title: "Practical Machine Learning - Prediction Assignment"
author: "attila.toth86"
output: html_document
---

# Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible 
to collect a large amount of data about personal activity relatively 
inexpensively. These type of devices are part of the quantified self movement – 
a group of enthusiasts who take measurements about themselves regularly to 
improve their health, to find patterns in their behavior, or because they are 
tech geeks. One thing that people regularly do is quantify how much of a 
particular activity they do, but they rarely quantify how well they do it. In 
this project, your goal will be to use data from accelerometers on the belt, 
forearm, arm, and dumbell of 6 participants. They were asked to perform barbell 
lifts correctly and incorrectly in 5 different ways. More information is 
available from the website here: http://groupware.les.inf.puc-rio.br/har (see 
the section on the Weight Lifting Exercise Dataset). 

# Data 

 - The training data for this project are available here: 
 https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
 - The test data are available here: 
 https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Let's download them if they are not yet in the working directory and load them 
into memory.

```{r cache=TRUE}
if(file.exists("pml-training.csv")==FALSE) 
    {
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
                  ,destfile = "pml-training.csv"
                  #,method="curl"
                  )
     }
if(file.exists("pml-testing.csv")==FALSE) 
    {
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
                  ,destfile = "pml-testing.csv"
                  #,method="curl"
                  )
     }
# Training dataset goes to variables with 'pml' prefix
pml <- read.csv("pml-training.csv")
```

The data for this project come from this source: 
http://groupware.les.inf.puc-rio.br/har.

# Goal

The goal of this project is to predict the manner in which they did the 
exercise. This is the `classe` variable in the training set. Any of the other 
variables are allowed use for prediction. This report describes how the model 
was built, how cross validation was used, what the expected out of sample error 
is, and why or how the choices were made.

# Preconditions

Let's load the necessary packages.

 - `caret` for machnine learning libraries
 - `plyr` & `dplyr` for data manipulation
 - `doParallel` for exploiting the power of multi core machines

```{r}
library(caret)
library(plyr)
library(dplyr)
library(doParallel)
```

I am "setting seed" for reproducible results.

```{r}
set.seed(1234)
```

# Data splitting

As always, my first step is to slice up our given 'testing' dataset to training 
and testing datasets that enables us to estimate out of sample error of a 
particular model. Testing dataset will not be touched until my model is ready.
Thus, any type of exploratory data analysis will be executed only on training 
set to avoid my model to be biased by testing data part. I used a randomly 
selected 75% of original training set to train my model.

```{r}
intrain <- createDataPartition(y = pml$classe, p=0.75, list=FALSE)
pmltrain <- pml[intrain,]
pmltest <- pml[-intrain,]
```

Let's take a look at the training dataset and the predictable outcome variable.

```{r}
str(pmltrain, list.len = 200)
summary(pmltrain$classe)
```

# Data Transformation and Processing

I decided to remove `user_name` since it can bring bias in the prediction 
(overfitting to this specific dataset) but would not add to the prediction power 
of a generalized model. (E.g. if Carlitos always performed his exercise 
correctly then learning algorythms would make use of this information and apply
this information for prediction. However, it might be useful information for our
given test set but not for generalized model, where we do not have the any user 
information.)

Also, removal of timestamp fields (i.e. `raw_timestamp_part_1`, 
`raw_timestamp_part_2`, `cvtd_timestamp`) makes sense, since these measures do
not add any value for a general model. In the time of this analysis meaning of 
variables of `new_window` & `num_window` were unknown for me but they clearly 
not reflect accelerometer measures so I left them out from the further analysis.
Variable of `X` has been ignored upfront as well, since it represents a simple
record identifier.

```{r}
pmltrainc <- pmltrain %>% select(-X,
                                 -user_name,
                                 -raw_timestamp_part_1,
                                 -raw_timestamp_part_2,
                                 -cvtd_timestamp,
                                 -new_window,
                                 -num_window,
                                 -classe)
```

I found that several factor variables were actually containing numeric
information. Just to make sure that a simple factor-numeric conversion would not
fail, I checked the levels for all factor variables.

```{r}
sapply(pmltrainc[,sapply(pmltrainc,is.factor)],levels)
```

Two particular values, `#DIV/0!` and empty values bring some data quality issue,
so I handled it by replacing them with `NA`.

```{r}
pmltrainc[pmltrainc=="#DIV/0!"] <- NA
pmltrainc[pmltrainc==""] <- NA
```

Then I converted all the factor variables to numeric.

```{r}
# http://stackoverflow.com/questions/8596466/r-change-all-columns-of-type-factor-to-numeric
# Applying the wisdom from Carl Witthoft above:
asNumeric <- function(x) as.numeric(as.character(x))
factorsNumeric <- function(d) modifyList(d, lapply(d[, sapply(d, is.factor)],   
                                                   asNumeric))

# Convert to numeric
pmltraincnum <- factorsNumeric(pmltrainc)
```

To avoid model building errors, I replaced NAs with zeros.

```{r}
pmltraincnum[is.na(pmltraincnum)] <- 0
```

Eventually I finalized the dataset for further analyses by including `classe`
target variable.

```{r}
pmltraincnum <- cbind(classe=pmltrain$classe,pmltraincnum)
```

These transformations have to be performed on test dataset as well.

```{r}
pmltestc <- pmltest %>% select(-X,
                               -user_name,
                               -raw_timestamp_part_1,
                               -raw_timestamp_part_2,
                               -cvtd_timestamp,
                               -new_window,
                               -num_window,
                               -classe)
pmltestc[pmltestc=="#DIV/0!"] <- NA
pmltestc[pmltestc==""] <- NA
pmltestcnum <- factorsNumeric(pmltestc)
pmltestcnum[is.na(pmltestcnum)] <- 0
pmltestcnum <- cbind(classe=pmltest$classe,pmltestcnum)
```

## Dimension Reduction

In order to build robust prediction model I needed to reduct the number of
independent variables. At this stage I still had 152 variables that is too much
for building a classifier model on top of that which returns results in a
reasonable time frame.

### Removing Zero Covariates

The first and most obvious task to be performed here is to identify and filter
out covariates with zero variance. (Since they will not have any explanation
power on my model for sure.)

```{r}
pmltraincnumnonzv <- pmltraincnum[,nearZeroVar(pmltraincnum,
                                               saveMetrics=TRUE)$zeroVar==FALSE]
```

The following 9 variables have been filter out in this step:

```{r}
rownames(
    nearZeroVar(pmltraincnum,
                saveMetrics=TRUE)[nearZeroVar(pmltraincnum,
                                              saveMetrics=TRUE)$zeroVar==TRUE,
                                  ]
        )
```

### Removing correlated predictors

In this section I considered to remove those predictors that have high (>0.99) 
correlation with one ore more other predictors, as they bring redundancy in the
model. (Correlation over 0.99 shows very high similarity between variables.)

```{r}
pmltraincnumnonzvnoncorr <- pmltraincnumnonzv[,-findCorrelation(cor(pmltraincnumnonzv[,-1]), cutoff = 0.99)]
```

This procedure removed the following 12 variables:

```{r}
colnames(pmltraincnumnonzv[,findCorrelation(cor(pmltraincnumnonzv[,-1]), cutoff = 0.99)])
```

### Removing Near Zero Covariates

I was still over 100 variables to include my model, so finally, I turned back to 
the previously used `nearZeroVar()` function to identify not only zero 
covariates but also **near** zero covariates.

```{r}
pmlfinaldata <- pmltraincnumnonzvnoncorr[,nearZeroVar(pmltraincnumnonzvnoncorr,saveMetrics = TRUE)$nzv==FALSE]
```

This is a longer list of variables that have been flagged as low contributors to
the total variability:

```{r}
rownames(
    nearZeroVar(pmltraincnumnonzvnoncorr,
                saveMetrics = TRUE)[nearZeroVar(pmltraincnumnonzvnoncorr,
                                                saveMetrics = TRUE)$nzv==TRUE,
                                    ]
        )
```

So I ended up with the following dataset to be analysed:

```{r}
str(pmlfinaldata)
```

# Model Fitting

First, let's train a model with Random Forest, as it is a widely used classifier. (I
could have used Linear Discriminants, multinomial logit, Support Vector 
Machines, etc..)

```{r cache=TRUE}
# Make use of process parallelization
registerDoParallel(makeCluster(detectCores()))

# Train the model
mdl <- train(classe~., 
             data=pmlfinaldata, 
             method = "rf", 
             trControl = trainControl(method = "cv", number = 5, allowParallel = TRUE)
             )
mdl
mdl$finalModel
```

The model accuracy is really good, over 99%. Let's take a look at the confusion
matrix:

```{r}
confusionMatrix(predict(mdl,pmltestcnum),pmltestcnum$classe)
```

According to model fitting results on the test set, I calculated the following
Out Of Sample Error rate of my model:

```{r}
prediction <- predict(mdl,pmltestcnum)
# Out of sample error percentage:
paste0(round(sum(pmltestcnum$classe != prediction)/length(prediction)*100,2),"%")
```

# Submission

For submission I read in the given test set that includes 20 observation.

```{r}
# Test dataset - for submission
orig_pml_test <- read.csv("pml-testing.csv")
```

I use my previously defined model to predict the requested `classe` variable.

```{r}
subm_pred <- predict(mdl,orig_pml_test)
subm_pred
```

I defined `pml_write_files` function to write out predictions into 20 separate 
files. (Source: Practical Machine Learning, Course Project Submission 
assignement instructions, 
url=https://class.coursera.org/predmachlearn-034/assignment/view?assignment_id=5)

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(subm_pred)
```

