---
title: "Reproducible Research - Peer Assessment 2"
author: "attila.toth86"
date: "20 Aug 2015"
output: html_document
---

# Synopsis

*This document is created in the framework of Reproducible Research online course provided by Johns Hopkins University on Coursera.*

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

The ultimate goal of this assignment is to explore the NOAA Storm Database and answer the following questions about severe weather events:

Accross the United States,
1) which types of events are most harmful with respect to population health?
2) which types of events have the greatest economic consequences?

# Data Processing

## Data description

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The data come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. Data is available for download [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

I first downloaded the NOAA database from course's website.

```{r}
if(!file.exists("StormData.csv.bz2")) {
    download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  destfile = "StormData.csv.bz2", method = "curl")    
}
```

There is some documentation of the database available. Here you will find how some of the variables are constructed/defined.

National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

After downloading I read in the given data from the file archive. read.csv() function can handle bzip files so no preliminary unzipping was required. (Due to significant filesize I check if the variable is already loaded to the environment in order to avoid unnecessary waiting time.)

```{r}
if(!exists("stormdata")){
    stormdata <- read.csv("StormData.csv.bz2")    
}
```

A high level overview of the database:
```{r}
str(stormdata)
```

For answering the given questions I used *EVTYPE*, *FATALITIES*, *INJURIES*, *PROPDMG*, *CROPDMG* variables.

EVTYPE variable gives us the observed and recorded event's type.
```{r}
summary(stormdata$EVTYPE)
```

FATALITIES variable gives us the number of recorded fatalities occured in a particular event.
```{r}
summary(stormdata$FATALITIES)
```

INJURIES variable gives us the number of recorded injuries occured in a particular event.
```{r}
summary(stormdata$INJURIES)
```

