---
title: "Tornados are the most harmful weather events for human health but floods cause the worst economic effect"
author: "attila.toth86"
date: "20 Aug 2015"
output: html_document
---

# Synopsis

*This document is created for the "Peer Assessment 2" assignment in the framework of Reproducible Research online course provided by Johns Hopkins University on Coursera.*

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

The ultimate goal of this assignment is to explore the NOAA Storm Database and answer the following questions about severe weather events:

Across the United States,

1) which types of events are most harmful with respect to population health?
2) which types of events have the greatest economic consequences?

# Data Processing

## Data Description

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The data come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. Data is available for download [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

There is some documentation of the database available. Here you will find how some of the variables are constructed/defined.

 - National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
 - National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Data Loading & Transformation

I first downloaded the NOAA database from course's website.
```{r}
if(!file.exists("StormData.csv.bz2")) {
    download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  destfile = "StormData.csv.bz2", method = "curl")    
}
```

After downloading I read in the given data from the file archive. read.csv() function can handle bzip files so no preliminary unzipping was required.
```{r cache=TRUE}
stormdata <- read.csv("StormData.csv.bz2")
```

A high level overview of the database shows that it consists of `r nrow(stormdata)` rows & `r ncol(stormdata)` columns (variables):
```{r}
str(stormdata)
```

From which my presumption is that answering the given questions following variables might be required: 

 - EVTYPE
 - FATALITIES
 - INJURIES
 - PROPDMG
 - PROPDMGEXP
 - CROPDMG
 - CROPDMGEXP

EVTYPE variable most likely gives us the observed and recorded events' type.
```{r}
summary(stormdata$EVTYPE)
```

FATALITIES & INJURIES seem to be numeric variables that indicate the number of recorded human fatalities/injuries occured in recorded events.
```{r}
summary(stormdata$FATALITIES)
summary(stormdata$INJURIES)

par(mfrow = c(2,2), mar = c(2,2,2,1))
hist(stormdata$FATALITIES, main = "Histogram of fatalities", ylim=c(0,80), xlab="")
hist(stormdata$INJURIES, main = "Histogram of injuries", ylim=c(0,150), xlab="")
boxplot(stormdata$FATALITIES, main = "Boxplot view of fatalities")
boxplot(stormdata$INJURIES, main = "Boxplot view of injuries")
```

Vast majority of recorded events do not have any effect on human's health.

PROPDMG (together with PROPDMGEXP) variable somehow refers to the value of property damage, CROPDMG (together with CROPDMGEXP) variable to the value of crop damage recorded in correspondence with a certain event.
```{r}
summary(stormdata$PROPDMG)
summary(stormdata$CROPDMG)

par(mfrow = c(2,2), mar = c(2,2,2,1))
hist(stormdata$PROPDMG, main = "Histogram of property damages", xlab="", ylim=c(0,15000))
hist(stormdata$CROPDMG, main = "Histogram of crop damages", xlab="", ylim=c(0,2000))
boxplot(stormdata$PROPDMG, main = "Boxplot view of property damages")
boxplot(stormdata$CROPDMG, main = "Boxplot view of crop damages")
```

Similarly to fatalities & injuries, most of recorded events do not have any economic effect.

PROPDMG & CROPDMG are numeric variables but alone they do not say much about damage extent they measure. For appropriate analysis I would need to know the unit that they are measured in. For this PROPDMGEXP & CROPDMGEXP variables seem to be the right choice despite the experienced data quality issues in these columns. Value set for these columns in the database:
```{r}
unique(stormdata$PROPDMGEXP)
unique(stormdata$CROPDMGEXP)
```

To narrow the dataset I have to work with I performed the following filters:

 - Selected only relevant columns: BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP
 - Filtered out those records that have no effect on human health & economy. (All relevant variables show zero.)
 
```{r results='hide', warning=FALSE}
library(dplyr)

sdata_n <- stormdata %>%
            filter(
                !(FATALITIES == 0 &
                INJURIES == 0 &
                PROPDMG == 0 &
                CROPDMG == 0)
                ) %>%
            select(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

Take a look at 'EXP' variables after filtering:
```{r}
sdata_n %>% group_by(PROPDMGEXP) %>% summarise(CNT=n())
sdata_n %>% group_by(CROPDMGEXP) %>% summarise(CNT=n())
```

There is no unambiguous information available about the coding so I first ignored the case of the letters then I assume the following value mappings applicable for these variables:

 - "H" = hundred USD
 - "K" = thousand USD
 - "M" = million USD
 - "B" = billion USD

The rest of the coding is unknown. I checked the extent of these values in light of their frequency & their correspondent value in the related variables (filtered out those records where the value field contains zero since there is no effect for those items whatever the 'EXP' columns' value set represents): 

```{r}
sdata_n %>% filter(PROPDMG!=0) %>% group_by(PROPDMGEXP) %>% summarise(CNT = n(),SUM = sum(PROPDMG)) %>% mutate(PERCENT_FREQ = round(CNT/sum(CNT)*100,2), PERCENT_VALUE = round(SUM/sum(SUM)*100,2))
sdata_n %>% filter(CROPDMG!=0) %>% group_by(CROPDMGEXP) %>% summarise(CNT = n(),SUM = sum(CROPDMG)) %>% mutate(PERCENT_FREQ = round(CNT/sum(CNT)*100,2), PERCENT_VALUE = round(SUM/sum(SUM)*100,2))
```

Looking at the frequency & extent of wrong codes, I believe, either exclusion or recoding would not affect my final result significantly, i.e. less the 1% is the occurence of unknown codes in the entire population. Therefore I decided to exclude these items from further analysis.

I translated the codes (H, K, M, B) to numeric format to use them as a multipicator of PROPDMG & CROPDMG variables in order to arrive to PROPDMGUSD & CROPDMGUSD variables that show the damage amount in a unified USD format throughout the entire database (e.g. PROPDMG = 25, PROPDMGEXP = K -> PROPDMGUSD = 25000). Problematic codes were simply translated to 0 (zero) so their multipicator effect will zero out the correspondent damage values.

For code translation I made use of *decode* function, created by [flodel](https://gist.github.com/flodel/4470993#file-decode-r).

Function description:
```{r}
decode <- function(x, search, replace, default = NULL) {
    decode.fun <- function(search, replace, default = NULL)
        if (length(search) == 0L) {
            function(x) if (is.null(default)) x else rep(default, length(x))
        } else {
            function(x) ifelse(x == search[1L], replace[1L],
                               decode.fun(tail(search,  -1L),
                                          tail(replace, -1L),
                                          default)(x))
        }
    return(decode.fun(search, replace, default)(x))
}
```

Code translation & some date formatting:
```{r}
sdata_clean  <- sdata_n %>%
                mutate(PROPDMGUSD = decode(toupper(PROPDMGEXP),
                                            c("K","M","B","H"),
                                            c(1000,1000000,1000000000,100),
                                            0)*PROPDMG,
                       CROPDMGUSD = decode(toupper(CROPDMGEXP),
                                            c("K","M","B","H"),
                                            c(1000,1000000,1000000000,100),
                                            0)*CROPDMG,
                       PROPDMGEXP = decode(toupper(PROPDMGEXP),
                                            c("K","M","B","H"),
                                            c("K","M","B","H"),
                                            NA),
                       CROPDMGEXP = decode(toupper(CROPDMGEXP),
                                            c("K","M","B","H"),
                                            c("K","M","B","H"),
                                            NA),
                       DATE = as.Date(as.character(BGN_DATE),"%m/%d/%Y")
                    ) %>%
                select(DATE, 
                       EVTYPE, 
                       FATALITIES, 
                       INJURIES, 
                       PROPDMG, 
                       PROPDMGEXP, 
                       PROPDMGUSD,
                       CROPDMG,
                       CROPDMGEXP,
                       CROPDMGUSD
                       )
str(sdata_clean)
```

# Results

## Most harmful events for population health

```{r echo=FALSE}
# Alter number display setup in document
options(scipen = 7)
```

In my analysis I consider those events harmful that resulted any fatality or injury. Since 1950 **in `r nrow(sdata_clean[sdata_clean$FATALITIES!=0,])` events** were reported that resulted one or more people's death. In these cases a total of **`r sum(sdata_clean$FATALITIES)` people deceased**. **Injuries** occured more often, **reported `r nrow(sdata_clean[sdata_clean$INJURIES!=0,])` times in which `r sum(sdata_clean$INJURIES)` people were involved.**

**Total number of events that have effect on human health counts `r nrow(sdata_clean[(sdata_clean$INJURIES!=0 | sdata_clean$FATALITIES!=0),])` in which `r sum(sdata_clean$FATALITIES) + sum(sdata_clean$INJURIES)` people were involved.**

I sumed fatalities & injuries by event type and listed them in a decreasing order of total number of people involved.

Top 10 events with regard to fatalities:
```{r}
evtype_fat_sum <-   sdata_clean %>%
                    group_by(EVTYPE) %>%
                    summarise(TOTAL_FATALITIES=sum(FATALITIES)) %>%
                    mutate(PERCENT = round(100*TOTAL_FATALITIES/sum(TOTAL_FATALITIES),2)) %>%
                    arrange(desc(TOTAL_FATALITIES))
evtype_fat_sum[1:10,]
```

Top 10 events with regard to injuries:
```{r}
evtype_inj_sum <-   sdata_clean %>%
                    group_by(EVTYPE) %>%
                    summarise(TOTAL_INJURIES=sum(INJURIES)) %>%
                    mutate(PERCENT = round(100*TOTAL_INJURIES/sum(TOTAL_INJURIES),2)) %>%
                    arrange(desc(TOTAL_INJURIES))
evtype_inj_sum[1:10,]
```

**What clearly seems from these numbers is that tornados can be considered as most dangerous weather events in respect of human health.**

## Events with greatest economic consequences

In order to answer the question of which weather event has the greatest economic consequences, I needed to create the measurement of economic effect. In my analysis this measure is simply the sum amount of damages occured in properties & crop together. Since 1950, **`r nrow(sdata_clean[sdata_clean$PROPDMGUSD!=0 | sdata_clean$CROPDMGUSD!=0,])` weather events were registered that resulted any economic damage**.

If I group the data by event type and sum the total damage (in property & crop) I will get to the conclusion that **floods (followed by hurricanes/typhoons) caused the most significant economic loss.**

Top 10 events in respect of economy damage:
```{r}
sdata_c_eco <-  sdata_clean %>%
                mutate(TOTAL_DMG_USD = PROPDMGUSD+CROPDMGUSD) %>%
                group_by(EVTYPE) %>%
                summarise(TOTDMG = sum(TOTAL_DMG_USD)) %>%
                mutate(PERCENT = round(100*TOTDMG/sum(TOTDMG),2)) %>%
                arrange(desc(TOTDMG))
sdata_c_eco[1:10,]
```
